Charged pion emission from neutron-rich heavy ion collisions for studies on the symmetry energy

### 2020-08-09 chapters_altFormat/ch_data_analysis.tex
Added Betty's edits

The S$\pi$RITROOT software is written in c++ and is based on the FAIRROOT package \cite{fairroot}. It is separated into modules which can be turned on or off depending on the particular analysis. While the base classes and frame work are provided by the FAIRROOT package, we added the specifics of the \spirit TPC and the algorithms for track reconstrucition. Here we will briefly describe the outline of the main tasks in the S$\pi$RITROOT software and then will discuss of each task later. The level of discussion concerning the software is really a practical, base understanding, appropriate for the discussion of this dissertation. For more details about the tracking algorithm the reader is refered to \cite{spiritroot_paper}, and to actual software which is published on GitHub \cite{spiritroot_git}.

@@

The fit of each individual pulse contains the height of the pulse and the timing information of the start of the pulse. We define the start of the pulse to be the location in time when the rising edge of the pulse reaches of 10\% of its peak value. The height of the pulse is related to the charge induced on the pad, $Q$. Combined with the start time of the pulse $t_o$  it is saved into a new container representing a \emph{hit}, called the \emph{STHit} class. After the PSA processes each pad in the raw event, an entire hit cloud of the event is produced. The x and z-coordinates come from the center of the pad which is already stored in the map of STPad. Since the STHit class inherits from the STPad class this information is also carried over. The y-coordinate is calculated using the known drift velocity $v$ and time of the signal $t_o$ as $y = v\cdot t_0$, $t_0$. The hit class forms the fundamental structure from which all tracking is performed since it contains the x,y,z spatial coordinates, and charge information $Q$.

@@

%\paragraph{Definition of clustering}
The hits within a helix track are dense and the position resolution of an individual hit is not very precise. As discussed in Section~\ref{sec:prf}, the TPC achieves sub-millimeter position resolution by combining neighboring pads into clusters using the the Pad Response Function (PRF). In the helix track finding task, the hits are further reduced into more precise clusters. A brief description of the method of clustering is illustrated in Figure \ref{fig:topview}. We have found that it is sufficient to simply cluster hits along one direction, either along the x or z-direction, depending on which direction is most perpendicular to the track. For example, the three clusters at the bottom of Figure \ref{fig:topview} are clustered along the x-axis, and the upper three are along the z-axis. The bold pads highlight the hits belonging to an example cluster from each set. Since we decide to cluster only in one direction, there is a natural inflection point in a track where the direction of clustering must switch. This switching point is determined by the crossing angle of the track $\theta$, which is defined as the angle between the tangent of the track, projected onto the pad plane, and the x-axis -- at a particular cluster location. In this definition, $\theta = \ang{90}$ corresponds to the case when the track is going along the z-axis and $\theta = \ang{0}$ for a track going along the x-axis.  The clustering direction in the case $\ang{45} < \theta \leq \ang{90} $ is along the x-direction. For tracks with angles $\ang{0} < \theta \leq \ang{45}$ we cluster along the z-axis.

@@

\subsection{Correction Task}
\label{sec:corrTask}
%extend dynamic range
%PRF cut
%space charge correciton
In this task the hit distribution, within a given cluster, is checked against the measured PRF discussed in Section~\ref{sec:prf}. Figure~\ref{fig:prfpimData} shows black lines around the measured PRF in the data representing gates defining the acceptable PRF region. If the charge distribution of the hits in a particular cluster is such that less than 50\% of the hits lie within this PRF gate, the cluster is thrown out from the track. Later, in Section~\ref{sec:expPRF}, we will describe how to define the PRF for a cluster. Clusters which do not follow this PRF gate usually have been corrupted in some way. Typically this occurs when charges from other nearby tracks mix distorting the hit distribution which corrupts the energy loss and position of the cluster. This tends to occur most frequently in the dense region surrounding the target, where the track density is high. But it may occur along the track for any other reason. The width of the gate is to allow for a reasonable range of PRF values to be accepted, without biasing the data too much.

@@

This task is repeated a second time, but this time including the vertex as a constraining point in which all tracks are refitted. In this way, the momentum resolution is greatly improved by adding the vertex as a constraining point; since the BDC detectors, described in Section~\ref{sec:bdc}, provide very accurate x and y values for the vertex location on the target. The energy loss of the track is calculated by the truncated mean method described in Eq.~\ref{eq:truncmean}. The GENFIT task represents the final part of the analysis which pertaining to track reconstruction. From here the PID can be constructed using the energy loss, momentum, and charge information.

\subsection{Vertex tracking (RAVE)}
\label{sec:vertex}
After all tracks are reconstructed by GENFIT, the tracks are then passed to the RAVE software package which reconstructs the event vertex from a collection of tracks \cite{rave}.  The RAVE vertex is typically referred to as the TPC vertex in this dissertation. It is used primarily to find if an event is on target by observing the z-distribution as will be discussed in Section~\ref{sec:vertexcut}. Off target events stemming from reactions in the active collimator or with the counter gas are rejected this way. After such off-target events are rejected,  we use the BDC vertex and refit the tracks as mentioned in the above section; this leads to more accurate momentum values than the TPC generated vertex provided by the RAVE routine.

@@

The gating grid noise profile was measured by recording experimental data, only turning off the anode wires. By doing so there are no signals originating from tracks in the TPC and there is only the gating grid noise, since the grid is still being triggered. The gating grid noise was stable for a given pad and the average value --shown as the red line-- can be calculated after averaging over several thousands of events and then removed by subtracting the profile from the real data for each pad.

The bottom panel of Fig.~\ref{fig:ggNoiseSub} shows the gating grid noise self correcting itself, representing a best case scenario. The pedestal -- or base line ADC-- is around 3800 ADC in the upper panel, is automatically subtracted when subtracting the gating grid noise profile since the subtracted ADC spectra is centered around 0. Still some residual gating grid noise remains even after the subtraction, but it is significantly reduced and its extent is over a much smaller time bucket region. This failure happens in regions where the derivative of the gating grid noise is large. In these areas the ADC value is dependent upon small jitters in the timing of the signal and as a result can vary significantly from event to event. To remove this residual noise in the subtracted spectra we ignored the first 30 time buckets of the readout.

@@

A cocktail of several light charged particles (p,d,t,${}^{3}$He,${}^{4}$He) of well defined magnetic rigidity was produced and measured in the TPC to provide a momentum calibration of the TPC. The magnetic and slit settings of the dipoles in the BigRIPS spectrometer were set so that the momentum resolution of the beam was defined to better than $\delta p/p$ < 1\%. Two magnetic rigidity settings were studied with no target in place. A thick Aluminum target was also inserted to provide a slightly lower energy calibration point for part of the lower rigidity setting, effectively creating three energy calibration points over several particle species. In some cases the certain particles were not able to propagate or were strongly attenuated and resulted in measurements of poor statistical precision.

Since the cocktail beam was a mono-energetic source of particles and rigidity selection in the BigRIPS corresponding to an actual momentum resolution of better than 1\%, we can interpret the  observed momentum resolution as a measurement of the intrinsic TPC momentum resolution of the TPC for these particles. The momentum resolution in general depends on several factors such as the particle's angle, momentum, charge, track multiplicity, etc. This calibration beam represents an ideal situation where the track was parallel to the pad plane and the track is not affected by the presence of other tracks. The average momentum resolution was measured to be $dp/p = 2\%$ over the full range of particle species for all settings as measured by the sigma of the Gaussian momentum distribution.

@@

Since the magnetic dipole setting of the BigRIPS spectrometer is well defined, we can calculate the expected momenta of each particle species measured. The energy losses through various materials in the beam line were propagated using LISE++ software \cite{lise++}. We found that the momenta of light particles in the  cocktail calibration beam differed significantly (up to 5\% ) from the expected values. This is shown in Tables~cref{tb:cocktail100tar,tb:cocktail100,tb:cocktail300}. We could address this discrepancy by incorporating the smaller horizontal magnetic field components at the edge of the dipole and taking into account the electron drift velocities in the direction of $\vec{E}\times\vec{B}$ as seen from Eq.\ref{eq:elecdrift}. The $\vec{E}\times\vec{B}$ drift velocity causes the electron trajectories to shift toward the +x-axis in the TPC coordinates causing particles of positive charge (going in the -x-axis) to have a higher measured momenta than in reality.  The details of the correcting for the $\vec{E}\times\vec{B}$ effect will be discussed later in Section~\ref{sec:spacecharge} in a more general way which also includes the correction for the presence of positive space charge in the field cage volume. The same correction technique was applied here, where the cocktail beam is the special case of zero space charge.

@@

\subsection{Anode gain calibration}
\label{sec:anodeCalib}
As discussed in Section~\ref{sec:wireplanes} the voltage of the anode sections 12 and 14 were reduced during the experiment, high currents being observed on the wires due to a leak in the gating grid that has since been corrected \cite{jon}. Out of all the runs used in the analysis in this thesis -- see \ref{tb:runList} -- the anode sections 12 and 14 were lowered to \SI{1085}{\volt} for runs 2272-2371 and set to \SI{1214}{\volt} for all other runs. By lowering the voltage on these anode wires, the gas gain is lowered as compared with all the other anode wire plane sections which operate at \SI{1460}{\volt}. To account for the drop in gain, we artificially increase the gain of the pads which lie above these anode wires in the software. The multiplicative factor is estimated by plotting the energy loss in the high gain sections relative to the low gain sections.

@@

Figure~\ref{fig:highlowcal} shows the correlation plot of $\langle dE/dx\rangle$ of the high versus the low gain sections. The effects of the high gain channel saturation around a value of 400 ADC\si{\per\milli\metre} and plateaus, whereas the low gain does not; this region is left out of the fit. A linear fit was performed for the calibration between the high gain channel $h_c = G\cdot l_c$, where $l_c$ is the value of the low gain channel and $G$ is the gain factor. In the case where the low anode voltage was \SI{1214}{\volt}, the gain factor was $G=9.8$.

@@

Using a TPC for measurements of HICs in intermediate energy heavy ion collisions presents a different set of challenges as opposed to higher energy experiments. Typically in higher energy experiments fundamental particles are produced with charge $\pm e$. Also, the particles are traveling at higher energies in which the energy losses are near or close to the minimum ionization. In these cases the dynamic range of most TPC electronics can cover a wide range of particles. In nuclear HICs, we are interested in measured particles with charges $Ze$ where Z=1-3, and even higher in some applications. Since the energy loss in the Bethe-Bloch equation, Eq.~\ref{eq:bb}, is proportional to $Z^2$, the range of energy losses reflects the possible range of z values. Even more, in low energy nuclear HIC of intermediate beam energies -- around \SI{300}{\MeVA} -- produce low velocity particles which exist in the $1/\beta^2$ region of Eq.~\ref{eq:bb}, where the energy loss grows dramatically. In this case, the limited dynamic range of electronics significantly limit the PID as the charge of a particle increases and the velocity decreases, leading to very large and even saturated pulses in the electronics.

Several TPCs have tried to address this issue by having regions of low and high gain, either in amplification gain or in electronics gain. This mitigates the complete loss of information but introduces a new problem. Particles which deposit large amounts of charge will have good measurements in the low gain areas, whereas particles depositing minimal energy losses will lose information in the same low gain areas. The reconstruction of such tracks will suffer. There are ongoing efforts in the nuclear community to develop new electronics which hope to mitigate these issues by developing more sophisticated  pre-amplifiers and electronics \cite{feanics,feanics2}.  Nevertheless, it is useful to develop a software technique which extends the dynamic range of TPC electronics without the use of new external hardware. Which can be particularly useful for experiments which have already taken data with older electronics technologies. In this section we will outline a novel software technique that takes advantage of the PRF described in Section~\ref{sec:prf} and can effectively extend the dynamic range of the TPC electronics.

In TPCs, the effective dynamic range is very different from the single channel dynamic range depending on how the TPC measurement is used. Typically TPCs are operated inside of a magnetic field for the purpose of reconstructing the momentum of a track, which requires sub-millimeter precision in the position determination along the track path. This is achieved by clustering several pads together as discussed in Section~\ref{sec:helixtrack}. To achieve this at least 2 adjacent pads must be measured, and the precision increases as the number of adjacent pads increases.

In this case the effective dynamic range is not the single channel dynamic range, but the relative dynamic range between central pad --holding the largest charge-- and the adjacent pads --holding the smaller charges in the PRF distribution. For example, to measure minimum ionizing particles, the Signal-to-Noise-Ratio (SNR) of the pad with the smallest charge in the distribution should be some reasonable value, so that the full charge distribution can be measured. From the PRF in Section~\ref{sec:prf}, we know the central pad in the cluster holds about 80\% of the total charge, whereas the two adjacent pads each hold the remaining 10\%. In the \spirit TPC the electronics gain was set so that the the pads have a SNR of 6:1 for minimum ionizing tracks, and therefore the central pad had a SNR of 50:1. The maximum SNR in the central pad -- before the electronics saturate -- the SNR is 800:1. Therefore the maximum SNR is roughly 16 times larger than that of minimum ionizing particles.

@@

The Gaussian fit was performed on each PRF for a crossing angle steps of  $\ang{5}$ ranging from $\ang{0} < \theta \leq \ang{90}$. Figure~\ref{fig:normsigma} shows the two parameters resulting from fitting the the Gaussian function -- Eq.~\ref{eq:prfgaus} -- which are plotted versus $\theta$. A $4^{th}$ order polynomial fit of these parameters allowed for interpolating for any given $\theta$ value, which is shown as the black line. Once we have this interpolation, we can predict the PRF for any given value of $\theta$.

@@

Figure~\ref{fig:lowvshigh} shows the the $\langle dE/dx\rangle$ values of the high gain region compared with the calibrated low gain region. The effect of saturation can be seen in the high gain region for the uncorrected data above values of \SI{400}{ADC \per \milli\metre} where the values plateau, whereas the low gain region still returns accurate values. Below this value the electronics are not saturated, and therefore the high and low gain sections agree. After applying the desaturation method, the correlation between the high and low gain sections is restored, as seen in Fig.~\ref{fig:lowvshigh}. From this comparison, we can approximate that the correction has corrected the high gain sections to agree with low gain sections to values of \SI{2000}{ADC \per \milli\metre}, increasing the dynamic range by a factor of a factor of at least 5.

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\linewidth]{spacechg_cartoon.png}
\caption{Location of space charge in 132 Sn}
\label{fig:spacechg_cartoon}
\end{figure}

@@

 The beam is positioned about \SI{25}{\centi\metre} below the anode plane and \SI{29.6}{\centi\metre} above the cathode in the TPC. It takes electrons approximately 5\si{\micro\sec} to drift to the anode plane whereas it takes the ions \num{5e4}\si{\micro\sec} to drift to the cathode. The beam rate in the experiment was approximately \SI{10}{\kilo\hertz}, which has an average occurrence of 1 beam every \SI{100}{\micro\sec}. This is  much shorter than the time it takes for the ions to terminate on the cathode plane, resulting in a build up of positive ions. Figure~\ref{fig:spacechg_cartoon} gives an idea of the shape of the sheet of space charge carved out by the beam path. The ions from each beam create a line charge which drifts towards the cathode with a constant velocity. The average distance between sequential ion paths is about \SI{25}{\micro\metre} apart, therefore we expect the average number of beam paths that make up the sheet charge is around \num{1440} tracks. Since the number of tracks in the sheet charge is large, and inter beam spacing is very small, we approximate the sheet charge as a uniform charge density.

@@

\begin{figure}[!htb]%
    \centering
    \subfloat[$\tin{132}{123}$ system.]{{\includegraphics[width=\textwidth,valign=c]{sn132_SC} \label{fig:132momdist_sc}}}%
    \qquad
    \subfloat[$\tin{108}{112}$ system.]{{\includegraphics[width=\textwidth,valign=c]{sn108_SC} \label{fig:108momdist_sc}}}%
	\caption{Momentum distributions of the Left and Right sides of the TPC for $\pi^+$ and $\pi^-$ particles.}
	\label{fig:sc_momdist}
\end{figure}

>>>>

The S$\pi$RITROOT software is written in c++ and is based on the FAIRROOT package \cite{fairroot}. It is separated into modules which can be turned on or off depending on the particular analysis. While the base classes and frame work are provided by the FAIRROOT package, we added the specifics of the \spirit TPC and the algorithms for track reconstrucition. Here we will briefly describe the outline of the main tasks in the S$\pi$RITROOT software and then will discuss of each task later. The level of discussion concerning the software is really a practical, base understanding, appropriate for the discussion of this Dissertation. For more details about the tracking algorithm the reader is refered to \cite{spiritroot_paper}, and to actual software which is published on GitHub \cite{spiritroot_git}.

@@

The fit of each individual pulse contains the height of the pulse and the timing information of the start of the pulse. We define the start of the pulse to be the location in time when the rising edge of the pulse reaches of 10\% of its peak value. The height of the pulse is related to the charge induced on the pad, $Q$. Combined with the start time of the pulse $t_o$  it is saved into a new container representing a \emph{hit}, called the \emph{STHit} class. After the PSA processes each pad in the raw event, an entire hit cloud of the event is produced. The x and z-coordinates come from the center of the pad which is already stored in the map of STPad. Since the STHit class inherits from the STPad class this information is also carried over. The y-coordinate is calculated using the known drift velocity $v$ and time of the signal $t_o$ as $y = v\cdot t_0$. The hit class forms the fundamental structure from which all tracking is performed since it contains the x,y,z spatial coordinates, and charge information $Q$.

@@

%\paragraph{Definition of clustering}
The hits within a helix track are dense and the position resolution of an individual hit is not very precise. As discussed in Section~\ref{sec:prf}, the TPC achieves sub-millimeter position resolution by combining neighboring pads into clusters using the Pad Response Function (PRF). In the helix track finding task, the hits are further reduced into more precise clusters. A brief description of the method of clustering is illustrated in Figure \ref{fig:topview}. We have found that it is sufficient to simply cluster hits along one direction, either along the x or z-direction, depending on which direction is most perpendicular to the track. For example, the three clusters at the bottom of Figure \ref{fig:topview} are clustered along the x-axis, and the upper three are along the z-axis. The bold pads highlight the hits belonging to an example cluster from each set. Since we decide to cluster only in one direction, there is a natural inflection point in a track where the direction of clustering must switch. This switching point is determined by the crossing angle of the track $\theta$, which is defined as the angle between the tangent of the track, projected onto the pad plane, and the x-axis -- at a particular cluster location. In this definition, $\theta = \ang{90}$ corresponds to the case when the track is going along the z-axis and $\theta = \ang{0}$ for a track going along the x-axis.  The clustering direction in the case $\ang{45} < \theta \leq \ang{90} $ is along the x-direction. For tracks with angles $\ang{0} < \theta \leq \ang{45}$ we cluster along the z-axis.

@@

\subsection{Correction Task}
\label{sec:corrTask}
%extend dynamic range
%PRF cut
%space charge correciton
In this task the hit distribution, within a given cluster, is checked against the measured PRF discussed in Section~\ref{sec:prf}. Figure~\ref{fig:prfpimData} shows black lines around the measured PRF in the data representing gates defining the acceptable PRF region. If the charge distribution of the hits in a particular cluster is such that less than 50\% of the hits lie within this PRF gate, the cluster is thrown out from the track. Later, in Section~\ref{sec:expPRF}, we will describe how to define the PRF for a cluster. Clusters which do not follow this PRF gate usually have been corrupted in some way. Typically this occurs when charges from other nearby tracks mix, distorting the hit distribution which corrupts the energy loss and position of the cluster. This tends to occur most frequently in the dense region surrounding the target, where the track density is high. But it may occur along the track for any other reason. The width of the gate is to allow for a reasonable range of PRF values to be accepted, without biasing the data too much.

@@

This task is repeated a second time, but this time including the vertex as a constraining point in which all tracks are refitted. In this way, the momentum resolution is greatly improved by adding the vertex as a constraining point; since the BDC detectors, described in Section~\ref{sec:bdc}, provide very accurate x and y values for the vertex location on the target. The energy loss of the track is calculated by the truncated mean method described in Eq.~\ref{eq:truncmean}. The GENFIT task represents the final part of the analysis pertaining to track reconstruction. From here the PID can be constructed using the energy loss, momentum, and charge information.

\subsection{Vertex tracking (RAVE)}
\label{sec:vertex}
After all tracks are reconstructed by GENFIT, the tracks are then passed to the RAVE software package which reconstructs the event vertex from a collection of tracks \cite{rave}.  The RAVE vertex is typically referred to as the TPC vertex in this dissertation. It is used primarily to find if an event is on target as will be discussed in Section~\ref{sec:vertexcut}. Off target events stemming from reactions in the active collimator or with the counter gas are rejected this way. After such off-target events are rejected,  we use the BDC vertex and refit the tracks as mentioned in the above section; this leads to more accurate momentum values than the TPC generated vertex provided by the RAVE routine.

@@

The gating grid noise profile was measured by recording experimental data, only turning off the anode wires. In doing so there are no signals originating from tracks in the TPC and there is only the gating grid noise, since the grid is still being triggered. The gating grid noise was stable for a given pad and the average value --shown as the red line-- can be calculated after averaging over several thousands of events and then removed by subtracting the profile from the real data for each pad.

The bottom panel of Fig.~\ref{fig:ggNoiseSub} shows the gating grid noise self correcting itself, representing a best case scenario. The pedestal -- or base line ADC-- is around 3800 ADC in the upper panel, is automatically subtracted when subtracting the gating grid noise profile since the subtracted ADC spectra is centered around 0. Still, some residual gating grid noise remains even after the subtraction, but it is significantly reduced and its extent is over a much smaller time bucket region. This failure happens in regions where the derivative of the gating grid noise is large. In these areas the ADC value is dependent upon small jitters in the timing of the signal and as a result can vary significantly from event to event. To remove this residual noise in the subtracted spectra we ignored the first 30 time buckets of the readout.

@@

A cocktail of several light charged particles (p,d,t,${}^{3}$He,${}^{4}$He) of well defined magnetic rigidity was produced and measured in the TPC to provide a momentum calibration of the TPC. The magnetic and slit settings of the dipoles in the BigRIPS spectrometer were set so that the momentum resolution of the beam was defined to better than $\delta p/p$ < 1\%. Two magnetic rigidity settings were studied with no target in place. A thick Aluminum target was also inserted to provide a slightly lower energy calibration point for part of the lower rigidity setting, effectively creating three energy calibration points over several particle species. In some cases certain particles were not able to propagate or were strongly attenuated and resulted in measurements of poor statistical precision.

Since the cocktail beam was a mono-energetic source of particles and rigidity selection in the BigRIPS corresponding to an actual momentum resolution of better than 1\%, we can interpret the  observed momentum resolution as a measurement of the intrinsic TPC momentum for these particles. The momentum resolution in general depends on several factors such as the particle's angle, momentum, charge, track multiplicity, etc. This calibration beam represents an ideal situation where the track was parallel to the pad plane and the track is not affected by the presence of other tracks. The average momentum resolution was measured to be $dp/p = 2\%$ over the full range of particle species for all settings as measured by the sigma of the Gaussian momentum distribution.

@@

Since the magnetic dipole setting of the BigRIPS spectrometer is well defined, we can calculate the expected momenta of each particle species measured. The energy losses through various materials in the beam line were propagated using LISE++ software \cite{lise++}. We found that the momenta of light particles in the  cocktail calibration beam differed significantly (up to 5\% ) from the expected values. This is shown in Tables~\cref{tb:cocktail100tar,tb:cocktail100,tb:cocktail300}. We could address this discrepancy by incorporating the smaller horizontal magnetic field components at the edge of the dipole and taking into account the electron drift velocities in the direction of $\vec{E}\times\vec{B}$ as seen from Eq.\ref{eq:elecdrift}. The $\vec{E}\times\vec{B}$ drift velocity causes the electron trajectories to shift toward the +x-axis in the TPC coordinates causing particles of positive charge (going in the -x-axis) to have a higher measured momenta than in reality.  The details of the correction for the $\vec{E}\times\vec{B}$ effect will be discussed later in Section~\ref{sec:spacecharge} in a more general way which also includes the correction for the presence of positive space charge in the field cage volume. The same correction technique was applied here, where the cocktail beam is the special case of zero space charge.

@@

\subsection{Anode gain calibration}
\label{sec:anodeCalib}
As discussed in Section~\ref{sec:wireplanes} the voltage of the anode sections 12 and 14 were reduced during the experiment as high currents were observed on the wires due to a leak in the gating grid that has since been corrected \cite{jon}. Out of all the runs used in the analysis in this thesis -- see Appendix~\ref{tb:runList} -- the anode sections 12 and 14 were lowered to \SI{1085}{\volt} for runs 2272-2371 and set to \SI{1214}{\volt} for all other runs. By lowering the voltage on these anode wires, the gas gain is lowered as compared with all the other anode wire plane sections which operate at \SI{1460}{\volt}. To account for the drop in gain, we artificially increase the gain of the pads which lie above these anode wires in the software. The multiplicative factor is estimated by plotting the energy loss in the high gain sections relative to the low gain sections.

@@

Figure~\ref{fig:highlowcal} shows the correlation plot of $\langle dE/dx\rangle$ of the high versus the low gain sections. The high gain channel saturates around a value of 400 ADC\si{\per\milli\metre} and plateaus, whereas the low gain does not; this region is left out of the fit. A linear fit was performed for the calibration between the high gain channel $h_c = G\cdot l_c$, where $l_c$ is the value of the low gain channel and $G$ is the gain factor. In the case where the low anode voltage was \SI{1214}{\volt}, the gain factor was $G=9.8$.

@@

Using a TPC for measurements of HICs in intermediate energy heavy ion collisions presents a different set of challenges as opposed to higher energy experiments. Typically in higher energy experiments fundamental particles are produced with charge $\pm e$. Also, the particles are traveling at higher energies in which the energy losses are near or close to the minimum ionization. In these cases the dynamic range of most TPC electronics can cover a wide range of particles. In nuclear HICs, we are interested in measured particles with charges $Ze$ where Z=1-3, and even higher in some applications. Since the energy loss in the Bethe-Bloch equation, Eq.~\ref{eq:bb}, is proportional to $Z^2$, the range of energy losses reflects the possible range of z values. HIC of intermediate beam energies -- around \SI{300}{\MeVA} -- produce low velocity particles which exist in the $1/\beta^2$ region of Eq.~\ref{eq:bb}, where the energy loss grows dramatically. In this case, the dynamic range of electronics significantly limit the PID as the charge of a particle increases and the velocity decreases, leading to very large and even saturated pulses in the electronics.

Several TPCs have tried to address this issue by having regions of low and high gain, either in amplification gain or in electronics gain. This mitigates the complete loss of information but introduces a new problem. Particles which deposit large amounts of charge will have good measurements in the low gain areas, whereas particles depositing minimal energy losses will lose information in the same low gain areas. The reconstruction of such tracks will suffer. There are ongoing efforts in the nuclear community to develop new electronics to mitigate these issues by developing more sophisticated  pre-amplifiers and electronics \cite{feanics,feanics2}.  Nevertheless, it is useful to develop a software technique which extends the dynamic range of TPC electronics without the use of new external hardware, which can be particularly useful for experiments which have already taken data with older electronics technologies. In this section we will outline a novel software technique that takes advantage of the PRF described in Section~\ref{sec:prf} and can effectively extend the dynamic range of the TPC electronics.

The effective dynamic range is very different from the single channel dynamic range depending on how the TPC measurement is used. Typically TPCs are operated inside of a magnetic field for the purpose of reconstructing the momentum of a track, which requires sub-millimeter precision in the position determination along the track path. This is achieved by clustering several pads together as discussed in Section~\ref{sec:helixtrack}. To achieve this at least 2 adjacent pads must be measured, and the precision increases as the number of adjacent pads increases.

In this case the effective dynamic range is not the single channel dynamic range, but the relative dynamic range between central pad --holding the largest charge-- and the adjacent pads --holding the smaller charges in the PRF distribution. For example, to measure minimum ionizing particles, the Signal-to-Noise-Ratio (SNR) of the pad with the smallest charge in the distribution should be some reasonable value, so that the full charge distribution can be measured. From the PRF in Section~\ref{sec:prf}, we know the central pad in the cluster holds about 80\% of the total charge, whereas the two adjacent pads each hold the remaining 10\%. In the \spirit TPC the electronics gain was set so that the pads have a SNR of 6:1 for minimum ionizing tracks, and therefore the central pad had a SNR of 50:1. The maximum SNR in the central pad -- before the electronics saturate -- the SNR is 800:1. Therefore the maximum SNR is roughly 16 times larger than that of minimum ionizing particles.

@@

The Gaussian fit was performed on each PRF for a crossing angle steps of  $\ang{5}$ ranging from $\ang{0} < \theta \leq \ang{90}$. Figure~\ref{fig:normsigma} shows the two parameters resulting from fitting the Gaussian function -- Eq.~\ref{eq:prfgaus} -- which are plotted versus $\theta$. A $4^{th}$ order polynomial fit of these parameters allowed for interpolating for any given $\theta$ value, which is shown as the black line. Once we have this interpolation, we can predict the PRF for any given value of $\theta$.

@@

Figure~\ref{fig:lowvshigh} shows the $\langle dE/dx\rangle$ values of the high gain region compared with the calibrated low gain region. The effect of saturation can be seen in the high gain region for the uncorrected data above values of \SI{400}{ADC \per \milli\metre} where the values plateau, whereas the low gain region still returns accurate values. Below this value the electronics are not saturated, and therefore the high and low gain sections agree. After applying the desaturation method, the correlation between the high and low gain sections is restored, as seen in Fig.~\ref{fig:lowvshigh}. From this comparison, we can approximate that the correction has corrected the high gain sections to agree with low gain sections to values of \SI{2000}{ADC \per \milli\metre}, increasing the dynamic range by a factor of a factor of at least 5.

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\linewidth]{spacechg_cartoon.png}
\caption{Cartoon diagram of the location of space charge for the ${}^{132}$Sn beam.}
\label{fig:spacechg_cartoon}
\end{figure}

@@

 The beam is positioned about \SI{25}{\centi\metre} below the anode plane and \SI{29.6}{\centi\metre} above the cathode in the TPC. It takes electrons approximately 5\si{\micro\sec} to drift to the anode plane whereas it takes the ions \num{5e4}\si{\micro\sec} to drift to the cathode. The beam rate in the experiment was approximately \SI{10}{\kilo\hertz}, which has an average occurrence of 1 beam every \SI{100}{\micro\sec}. This is  much shorter than the time it takes for the ions to terminate on the cathode plane, resulting in a build up of positive ions. Figure~\ref{fig:spacechg_cartoon} gives an idea of the shape of the sheet of space charge carved out by the beam path. Figure~\ref{fig:beampaths} show the beam paths of all beams as extracted from the data. The ions from each beam create a line charge which drifts towards the cathode with a constant velocity. The average distance between sequential ion paths is about \SI{25}{\micro\metre} apart, therefore we expect the average number of beam paths that make up the sheet charge is around \num{1440} tracks. Since the number of tracks in the sheet charge is large, and inter beam spacing is very small, we approximate the sheet charge as a uniform charge density.

@@

\begin{figure}[!htb]%
    \centering
    \subfloat[$\tin{132}{124}$ system.]{{\includegraphics[width=\textwidth,valign=c]{sn132_SC} \label{fig:132momdist_sc}}}%
    \qquad
    \subfloat[$\tin{108}{112}$ system.]{{\includegraphics[width=\textwidth,valign=c]{sn108_SC} \label{fig:108momdist_sc}}}%
	\caption{Momentum distributions of the Left and Right sides of the TPC for $\pi^+$ and $\pi^-$ particles.}
	\label{fig:sc_momdist}
\end{figure}
